\section{Diseases}

\begin{multicols}{2}

Fenestra's illneses, and cures for them.

\subsubsection{Breathrot}

One's breath becomes foul, as thier lungs slowly fill with stagnant air. One affected can never have less than 1 Fatigue. Every interval this effect increases by 1.

\subsubsection{Corpse Hands}

Those unfortunate to ail from this will see first their fingertips, then their hands, and finally whole arms, dry up and become stiff. Gain -1 penalty to Dexterity. Every 3 intervals effect increases by 1.

\subsubsection{Mindflash Syndrome}

Sufferer's mind becomes restless, keeping constant watch and clouding judgement. -1 Intelligence, but +1 Wits, no sleep is possible. Every interval Intelligence worsens by 1, and every 2 intervals Wits increases by 1. Paranoia and hallucinations are not uncommon after a while.

\subsubsection{Hero's Bane}

Your legs and arms feel too heavy to move. -1 Strength and Speed. Every interval, the higher of the two gains additional -1 penalty, and if they are equal, Strength penalty increases by 1.

\subsubsection{Woodspy's Choke}

Your throat feels so tight that you can only whisper. Talking loudly inflicts so much pain, that you suffer +1 Fatigue.

\subsubsection{Torpid Flesh}

If you spend more than 3 AP in a turn, gain 1 Fatigue. Same if you perform any Strength or Speed check.

\end{multicols}

\section{Poisons}

\begin{multicols}{2}

Poisons available for crafting and purchase.

\subsubsection{Downroot Poison}

-1 to all attributes for one day. Extracted from Downroot.

\subsubsection{Magebane}

One afflicted with it can't regenerate MP at the end of this interval. +1 difficulty to cast all spells. Extracted from Mage Oak.

\subsubsection{Assasin's Mark}

Whenever you're in dark, you start to faintly glow. If you spend most of the interval like this, gain 1 Fatigue. Extracted from Glowshroom. Assasins use this poison to mark their prey, so they can find them easily.

\end{multicols}
