\documentclass[10pt,twoside]{book}
\usepackage{config/bind}
\usepackage{config/booklet}
\setcounter{bookLevel}{2}
\externalReferent{core}
\externalReferent{stories}
\externalReferent{judgement}

\begin{document}

\frontmatter

\miniCover{Faces}{
  \par in the \Glsfmttext{deep}
}

\noindent
Need some \glspl{npc} for the \gls{deep}?
Here's a collection of \glspl{statblock}.

\clearpage
\pagestyle{minizine}

\subsection{Trade in the \Glsfmttext{deep}}

Dwarven traders move small produce in a backpack, and larger produce in a hand-cart.

\dwarventrader

\dwarventrader

\stepcounter{treasure}

\begin{dlistDouble}
  \item Have this \gls{talisman}: \lootTalisman.
  Nearly free.
  Just looking for \arabic{r3}0 goblin heads.

  \showTalisman
  \item Have 4 barrels of spirits, want a proper tale from the surface world (3 hours, minimum).
  \item Have \lootJewellery, want \glspl{coin}.
  \item Have contracted Mindflash syndrome, and want a cure.
  \item Have $1D6$ alcohol-lanterns, want 8~\glspl{sp} each for them.
  \item Have information on how to leave, want food.
  \item Have $1D6$ shortswords and $1D6$ coils of rope (15~\glspl{step} long), want novel weapons.
  \item Have chainmail of all sizes, want to hear an oath to protect the mountain at all cost and any call.
  \item Have beard-dyes (\textit{all the colours of the furnace!}), want 2~\glspl{gp} per phial.
  \item Have this idea for making money by selling in a non-transitive circle.
  \textit{It'll make money from nothing, honestly.
  Just gimme some coin and return in a year or so.}
  \item Wedding invitations to be delivered to the next settlement, want to know how much it'll cost if you carry them all.
\end{dlistDouble}

Dwarven traders often refuse to sell to people who don't know how to use what they're buying.
Waste is worse than theft.

\dwarventrader

\subsection{Dwarven Plans}

\begin{dlist}
  \item Find more birdies for the mine --- we only have two\ldots ONE left.
  \item Making a model stone bridge out of pebbles.
  \item Discussing how to test how tasy an ooze is befor eating it.
  \item Competitive pipe-smoking.
  \item Singing down a deep hole (the echo is part of the song).
  \item Pleating beards.
\end{dlist}

\dwarvensoldier

\dwarvensoldier

\dwarvensoldier

\dwarvenrunemaster

\showStdSpells

\clearpage

\subsection{The Dead in the \Glsfmttext{deep}}

\Glsfmtname{undead} \glsentrydesc{undead}

\ifnum\thepage<9
  \ghoul
\fi

\ghoul

\ghast

\ghast

\clearpage

\demilich

\showStdSpells

\ifnum\thepage<12
  \ghast

  \ghast
\fi

\clearpage

\demilich

\showStdSpells

\clearpage

\ifnum\thepage<13
  \undeadgoblin

  \undeadhobgoblin
\fi


\ifnum\thepage<14
  \clearpage
  \undeadogre
\fi

\end{document}
