\documentclass[10pt,twoside]{book}
\usepackage{config/bind}
\usepackage{config/booklet}
\setcounter{bookLevel}{2}
\externalReferent{core}
\externalReferent{stories}
\externalReferent{judgement}

\begin{document}

\frontmatter

\miniCover{Faces}{
  \par in the warren
}

\noindent
Need some \glspl{npc} for a warren?
Here's a collection of \glspl{statblock}.

\clearpage
\pagestyle{minizine}

\subsection{Gnomish Plans}

\begin{dlist}
  \item Discussing a rock-hiding game (it's only \emph{played} in the dark).
  \item Drawing schematics for a spike-pit-trap on the ground with a stick.
  \item Discussing next year's carrot harvest.
  \item Competitive pipe-smoking.
  \item Discussing how to return home with their enormous clay-jar-trap.
  It's as tall as a human, and full to the brim with an ooze!
  \item Walking slowly, while tapping every bit of ground with a walking stick
  (there is no trap).
\end{dlist}

\jelly

\gnome

\gnome

\gnomishsoldier

\gnomishsoldier

\gnomishillusionist

\showStdSpells

\gnomishillusionist

\showStdSpells

\subsection{Goblin Ideas}

\begin{dlist}
  \item A night on the town! (or \gls{village})
  Let's race!
  \item {\large \underline{Screeching}} competition!
  \item All sleep with a gentle snore like a distant earthquake.
  \ifnum\thetemperature=0%
    \item Let's build sleds!
    \item Let's use the sleds!
    \item Half sleep silently while the rest spit snot at the fire to hear it crackle.
  \else%
    \item Let's imitate birdies to see if it lures them down.
    \item Blind the thief!
    Stab his eyes out!
    He stole my stuff!
    {\sffamily (\lootGoblin)}
    \item Is it moral to eat human babies when they haven't even tried to kill us yet?
  \fi%
\end{dlist}

\goblin

\goblin

\goblin[\NPC{\N}% Symbol
  {\ifcase\value{r2b} Hunchbacked \or Half-toothless \else Sweaty\fi~Goblin}% Name
  {\ifcase\theStrength%
    Pustule-covered
  \else%
    Black-spotted
  \fi
  \ifcase\theSpeed%
    feet%
  \or%
    ears
  \else%
    face%
  \fi%
  }% Description
  {\ifodd\value{r3} scratches \else rubs \fi \ifodd\value{r2} eyeballs\else chest\fi}% Mannerism
  {\ifodd\thexp to eat snails\else to collect rocks\fi}% Wants
  ]

\goblin[\NPC{\N}% Symbol
  {\ifcase\value{r2b} Hunchbacked \or Half-toothless \else Sweaty\fi~Goblin}% Name
  {\ifcase\theStrength%
    Pustule-covered
  \else%
    Black-spotted
  \fi
  \ifcase\theSpeed%
    feet%
  \or%
    ears
  \else%
    face%
  \fi%
  }% Description
  {\ifodd\value{r3} scratches \else rubs \fi \ifodd\value{r2} eyeballs \else chest\fi}% Mannerism
  {\ifodd\thexp to eat snails\else to collect rocks\fi}% Wants
  ]

\wolf

\goblincaster

\showStdSpells

\ifnum\theWater>1
  \morphwolf
\fi

\hobgoblin[\NPC{\T[\arabic{r12}]\N}% Symbol
  {\ifodd\theSpeed Stinking \else Vomit-Stained\fi~Hobgoblins}% Name
  {\ifnum\theDexterity<-1%
    Flaking, ragged,
  \else%
    Hairy
  \fi
  \ifcase\theWits%
    toenails%
  \or%
    ear-tips%
  \else%
    face-scabs%
  \fi%
  \ifodd\value{r3}\else, missing eye\fi%
  }% Description
  {\ifodd\value{r3} chews on own \else gently massages \fi \ifodd\value{r2} hands \else tongue\fi}% Mannerism
  {\ifcase\value{r3} to find somewhere {\ifnum\thetemperature=0 warm \else to sleep \fi}\or to eat dogs\else sharper teeth\fi}% Wants
]

\hobgoblin[\NPC{\ifodd\theStrength\M\else\F\fi\N}% Symbol
  {\ifodd\theSpeed Stinking \else Vomit-Stained\fi~Hobgoblin}% Name
  {\ifnum\theDexterity<-1%
    Flaking, ragged,
  \else%
    Hairy
  \fi
  \ifcase\theWits%
    toenails%
  \or%
    ear-tips%
  \else%
    face-scabs%
  \fi%
  \ifodd\value{r3}\else, missing eye\fi%
  }% Description
  {\ifodd\value{r3} chews on own \else gently massages \fi \ifodd\value{r2} hands \else tongue\fi}% Mannerism
  {\ifcase\value{r3} to find somewhere {\ifnum\thetemperature=0 warm \else to sleep \fi}\or to eat dogs\else sharper teeth\fi}% Wants
]

\ogre

\ogre

\armouredOgre

\armouredOgre

\ifnum\thetemperature>1
  \hulk
  \hulk
\else
  \watcher
  \jelly
\fi

\clearpage

\subsection{Things Lying on the Ground}

\setcounter{goblinLoot}{6}
\addtocounter{goblinLoot}{\value{r6}}
\begin{dlist}
  \item \lootGoblin
  \item \lootGoblin
  \item \lootGoblin
  \item \lootGoblin
  \item \lootGoblin
  \item \lootGoblin
\end{dlist}

\ifnum\thepage<16
  \newpage
  \jelly
\fi

\end{document}
