\subsubsection{Millbroch}
processes the flour for the surrounding \glspl{village} with the massive windmill on the side.
\ifcase\thetemperature%
  But not today, as these days are for snow, and hiding from the open air.
  The air up this hill can pull the heat from your bones, and every night \glspl{griffin} fly overhead, scouting hungrily, and occasionally imitating the pipers.

  \griffin
  \ifnum\thepage<14 \griffin\fi

  \Glspl{griffin} like to swoop down in the dark, so anyone who goes out at night takes \pgls{dawnDolly} on their back.
  The \gls{broch}'s lower floor has an assembly of \glspl{dawnDolly}, mostly face-down on the dust-white ground, as if hiding from the prospect of going outside.
  Bits and pieces of the straw-people lie scattered around the \gls{broch} after a disappointed \gls{griffin} tore its limbs apart mid-flight.

  The \glspl{dawnDolly} come from the generous \gls{templeOfPoison}, who take a cut from every nearby \gls{village} when processing their wheat.
\or%
  But not today.

  \IfLabelExistsTF{\jobname:harvestCrawler}{
    \chitincrawler

    \begin{boxtext}
      Approaching the famous Millbroch, you can see dirty blades spinning slowly in the wind.
      As a blade rises, something on it wriggles, it cries out for help.
      Then the dark centre stretches legs, and in one single stride jumps up from the middle of the mill, to the top, and the voice cries out in pain.
    \end{boxtext}

    \Gls{cutter} \composeHumanName\ won't shout again, as he knows it attracts the \gls{crawler} to deposit more webbing.
    He just bleeds silently from the creature's knife-like appendages.
  }{
    \begin{boxtext}
      Approaching the famous Millbroch, you can see dirty blades have stopped, stuck.
    \end{boxtext}

    \Pgls{crawler} has left its webbing on the mill, then left out of irritation with the moving blades.
    The \gls{crawler} has left hours ago, but the blades remain stuck.
  }

\or%
  \Glspl{ranger} always avoid the \gls{broch}, complaining that they have better things to do than `militant milling'.

  \begin{speechtext}
    If another piper dies because something crawled up those blades, there will be \emph{\underline{words}}.
  \end{speechtext}

  They say `words', but of course the \glspl{ranger} plan to mutiny, and kill \Gls{jotter}~\composeHumanName\ if anyone else dies just to keep his cousin in the \gls{templeOfPoison} happy.

  Anyone of lower rank will find the \gls{broch} full of work, as \glspl{gDigger} maintain the machinery, \glspl{soldier} go back and forth to every nearby \gls{village}, and \glspl{cutter} scout the forests for miles around for \glspl{monster} of any type.
\else%
  \Gls{landlord}~\marketInnOne, from the \gls{templeOfPoison} visits occasionally, as she sponsors the `little experiment in cross-\gls{temple} relationships'.
  Today, she's arrived with a caravan, with more barrels of ale than the \gls{broch} has ever seen.
  This is good for morale, but bad for the mill, as the \glspl{guard} here, constantly rotating and dying, have never been good at the job of milling, and don't properly understand fire-safety procedures.

  Tonight, the grain runs empty, and the mill just keeps grinding.
  Little sparks ignite the dusty flour everywhere, a fire begins.

  The troupe can roll \roll{Intelligence}{Vigilance} at \tn[10] to understand the danger before it begins.
  If that fails, they can roll \roll{Wits}{Empathy} to rally the drunken \glspl{guard} to put the fire out at \tn[12] (the \gls{broch} has drunk its ale, and has very little water).
\fi%

