\ifcase\value{cycle}

% Yonders

\stepDay{2}

\begin{boxtext}
  Last night, the rain started.
  This morning, it continued.
  Buckets of rain, so loud it's hard to speak.
  Everything feels heavy, and the road has started to turn into a bog.
\end{boxtext}

A long head rears over the forest's canopy in the distance.
A \roll{Wits}{Vigilance} roll at \tn[10] lets the \glspl{pc} spot the \gls{basilisk}, with three \glspl{round} to prepare.

Otherwise, they hear it coming when it's twenty~\glspl{step} away, slinking through the forest, and charges them.

\basilisk

\or% Sables

\stepDay{3}

\begin{boxtext}
  Tonight, the biting winds carry wolf-howls.
\end{boxtext}

The wolves will investigate the troupe, and see if they can steal some food, but leave if threatened.

\wolf

\or% Wrecans

\stepDay{3}

A \roll{Wits}{Survival} roll at \tn[10] lets the \glspl{pc} spot boar-tracks.
He's close, and they can find him half a mile upwind.

\boar

Sneaking up just needs a \roll{Dexterity}{Stealth} roll at
\setTN{Wits}{Vigilance}
\tn.

\or% Abderians

\stepDay{3}

\begin{boxtext}
  \Pgls{crawler} stands in the road, flailing its arms madly.
\end{boxtext}

A second \gls{crawler} lies nearby, waiting to pick off people at the back.
A \roll{Intelligence}{Survival} roll at \tn[10] can tell the troupe this is commong during mating season.

\chitincrawler

\chitincrawler

\or% Sylfs

\stepDay{2}

\begin{boxtext}
  A massive web lies across the road ahead.
  The trees to the left look rancid with old and new webs.
  Forty \glspl{step} behind, a brown thorax pokes out of the forest and begins making a squelching sound as webbing shoots out and sticks to the trees.
  The hairy thorax wipes the webbinng onto a tree-trunk, completing the web across the road, then walks into the forest.
\end{boxtext}

The forest to the troupe's right has no webs, just \pgls{crawler}.
Two \glspl{crawler} made this mass of webbing, but the other is away, hunting.

This windless \gls{interval} brings only 2~\glspl{mp}.

\else% Eldrens

\stepDay{5}

\begin{boxtext}
  Sunrise is cancelled.
  Instead, the morning slowly reveals mist as thick as clouds.
\end{boxtext}

\Glspl{griffin} like using the mist to hunt.
First one dive-bombs the troupe, trying to pick one up, or just strike it and jump back.
Then another.

\griffin

The troupe can see the first coming with a \roll{Wits}{Vigilance} roll at \tn[12].
Each Failure Margin removes \pgls{ap}, and three dive within the first \glspl{round}.
They attack the tallest person first, then any horses present.

If one dies, the rest retreat and watch from the trees, occasionally mimicking anything they hear, and wait to attack again at night.

\fi
