\ifnum\value{cycle}=0

\stepDay{2}
\Pgls{woodspy} hangs from a tree, above the road, and the clear blue sky makes it particularly easy to spot.
The troupe can roll \roll{Wits}{Vigilance} at \tn[10].
Each Success Margin gives them 10~\glspl{step} distance, while failure removes \pgls{ap}.

\woodspy

If the \gls{woodspy} dies, it turns white with rage, then fades to brown as its blue blood drains.

\else

\basilisk

\stepDay{2}

\ifnum\theAthletics>1
  \begin{boxtext}
  Trees in the distance bend outward.
  Green legs slink up further, with a tail wrapped below for balance.
  And as the monster watches you, branches the size of tent-poles begin to break under its weight.
  It drops below the canopy, out of sight.
  \end{boxtext}
\else
  \begin{boxtext}
    Three puddles lie connected.
    They form the splayed toes of a bird, if birds had feet larger than your head.
  \end{boxtext}

  \Pgls{basilisk} left the footprint-puddle yesterday.

  \paragraph{If the troupe visit any local \glspl{village},}
  they find people have become paranoid, as they suffer from Mindflash SyndromeMindflash Syndrome.%
  \exRef{judgement}{Judgement}{Mindflash Syndrome}
  Something upriver is making them sick, though the farmers argue about the cause.

  \paragraph{If the troupe track the \gls{basilisk},}
  they find it five miles upriver, inside the river, where nothing can smell it.
  The \gls{basilisk} raises its nostrils above the surface every few minutes.
  The troupe can try a \roll{Wits}{Vigilance} roll at \tn[10] to spot the \gls{ambush}.

\fi


\fi

