\ifnum\value{cycle}=1

\stepDay{3}
\begin{boxtext}
  Snow dances down.
  The morning sparkles.
  A massive web stands frozen between two trees nearby.
\end{boxtext}

If the troupe search around, they may find \arabic{r2t4} hibernating \glspl{crawler} under a pile of snow (\roll{Intelligence}{Survival}, \tn[10]).
Any harsh movements (stamping on the ground, or hitting them) awakens them.
They wake with a -8~Penalty, which decreases by 1 each \gls{round}.

\IfLabelExistsTF{crawlerStatblock}{
  Use the stats \vpageref{\jobname:harvestCrawler}.
}{
  \chitincrawler[\npc{\T[4]\A}{Hibernating \Glsfmtplural{crawler}}]
}

\else

\ifnum\value{r12}=8

\stepDay{1}
\Pgls{pc} armour has become ragged with age, and needs some better care.

\begin{boxtext}
  Your leather armour's getting old.
  The straps look worn out, and pieces have begun flaking off as you walk.
\end{boxtext}

Which \gls{pc} this affects, depends on a few factors.

\begin{enumerate}
  \item
  Start with the \glspl{armour} with the highest \gls{dr}.
  \item
  Out of those, look at the \gls{pc} with the lowest \roll{Intelligence}{Crafts} total.
  \item
  If multiple \glspl{pc} have the same Bonus, select whoever has the most \glspl{fp}.
\end{enumerate}

Taking care of the \gls{armour} requires an \roll{Intelligence}{Crafts} roll at \tn[10] and a workshop.
Until then, each time the \gls{pc} moves quickly (sprinting, fighting, et c.) the armour has a 1-in-6 chance of something snapping, making it open in all the wrong places, and flail about.
When this happens, the \gls{armour}'s \gls{covering} decreases by~2, and the \gls{pc} takes a -2~Penalty to Speed as straps fall around their legs or a breastplate starts banging into their legs.

\fi

\fi
