\ifnum\value{cycle}>4

  \stepDay{4}
  \ifnum\value{r12}>9%
    A pregnant \gls{crawler} sets an ambush, then waits.

    \chitincrawler[\npc{\F\A}{Pregnant \Glsfmttext{crawler}}]
    \label{crawlerStatblock}

    \begin{boxtext}
      A long, thick web sparkles in the Sun, stretched across the road, between two trees, twenty \glspl{step} away.
      The horses stop, and try to pull to the side.
    \end{boxtext}

    If the \glspl{pc} kill her and harvest the body, the eggs hatch and begin eating her from the inside.
    This raises the \gls{harvesting} \gls{tn} by~2.
    However, on success, \pgls{pc} can kill the swarm as they hatch, one by one, while preserving the value of the corpse.

    \crawlerSwarm

  \else%
    \begin{boxtext}
      Clear skies send rods of Sunlight down through the black forest.
      The rare glimpse into the blackness mostly shows tree-bark, but one beam illuminates a shrivelled sack, like an old sleeping back or like a giant walnut, around a hundred \glspl{step} away.
    \end{boxtext}

    The Sunbeam has hit \pgls{bedshroom}.
    They \glsentrydesc{bedshroom}.

  \fi

\else
  \ifnum\value{temperature}=1
    \stepDay{4}
    \Glspl{griffin} love to hunt on overcast days.
    Spotting it requires a \roll{Wits}{Vigilance} roll at \tn[12], and each Failure Margin removes \pgls{ap}.

    \griffin[\npc{\A}{\Glsentrytext{griffin}}]

    The \gls{griffin} dives down to the smallest member of the troupe (perhaps an animal) and tries to grab them, then fly off in a single dive-bomb motion.
    If it fails, it flees.

  \fi

\fi

