\ifnum\value{cycle}=3

\ifodd\value{r2}

  \stepDay{2}

  \begin{boxtext}
    The rain smashes the ground, pulverising it to mud.
    And over the sound of rain, something screams in pain.
  \end{boxtext}

  The \glspl{pc} can try an \roll{Intelligence}{Survival} roll at \tn[10] to understand this sound comes from a deer.
  If they beat \tn[12], they figure out that deer usually make this sound when \pgls{digger} catches them.
  \Glspl{digger} have no voice, so their `mating call' consists of catching an animal, pulling it half-way into their warren, and letting it scream until it dies.

  If the troupe investigate, they find a deer, stuck half-way down \pgls{digger} tunnel.
  Within \pgls{interval}, a female \gls{digger} arrives, but runs away if it spots the \glspl{pc}.

\else

  \stepDay{1}

  \begin{boxtext}
    The wind howls.
    Something else howls too, but it's hard to hear over the wind.
    Both sound like they're in pain.
  \end{boxtext}

  The \glspl{pc} can try an \roll{Intelligence}{Survival} roll at \tn[10] to understand this sound comes from a wolf.
  If they beat \tn[12], they figure out that wolves usually make this sound when \pgls{digger} catches them.
  \Glspl{digger} have no voice, so their `mating call' consists of catching an animal, pulling it half-way into their warren, and letting it scream until it dies.

  If the troupe investigate, they find the wolf stuck half-way down \pgls{digger} tunnel.
  Within \pgls{interval}, a female \gls{digger} arrives, but runs away if it spots the \glspl{pc}.

\fi

\fi
